\newpage 

```{r dependencies, echo = FALSE, message = FALSE, warning = FALSE}

library(brms)
library(dplyr)
library(purrr)
library(ggplot2)
library(grid)
library(gridExtra)
library(Rcompadre)
library(patchwork)
library(Rpadrino)
library(knitr)
library(kableExtra)
library(sf)
library(png)
library(BIEN)
library(mgcv)
library(tools)


```

# Summary 

The study of structured populations dynamics has become a key aspect of many sub-disciplines within ecology as a whole. Structured population models provide an elegant tool to scale from individual observations to population level inference. Life tables are probably the most widely used tool to study human demography. Currently, matrix population models (MPMs) are the most widely used structured population models in ecology. Integral projection models (IPMs) are newer, but are seeing increasingly frequent use in the literature. IPMs allow researchers to study the demography of population structured y one or more continuously distributed traits in discrete time. Concurrently, researchers have developed a variety of tools and techniques to parameterize and implement IPMs. However, pressing questions in ecology, evolution, and conservation, and their associated data and computational requirements, have outpaced the development of tooling to aid their implementation, interpretation, and synthesis. This dissertation contributes to filling these gaps by introducing new computational tools and data collection techniques, and then applying them them to an invasive plant genus that has invaded multiple continents. These tools and analyses enable large scale data collection, ease the implementation of complex IPMs, and provide a more streamlined experience for syntheses.

In **Chapter 2**, I introduce a new _R_ package, `ipmr`, to implement and analyse IPMs. This _R_ package provides a domain specific language to implement deterministic and stochastic models, with two options for handling the latter. Furthermore, `ipmr` handles density dependence with minimal additional overhead. Density dependence is a critically important population process that, until `ipmr`, was not implemented in any _R_ packages for structured populations. `ipmr` is designed to accommodate almost any type of regression model used in the vital rate fitting process, and also includes methods for extracting the basic building blocks used in many downstream analyses. This chapter includes two appendices with detailed case studies demonstrating the package's use, and the package itself contains six vignettes that are intended for audiences from a range of technical backgrounds. 

In **Chapter 3**, I introduce two additional tools: the PADRINO IPM Database, and `Rpadrino`, an _R_ package for interfacing with PADRINO. PADRINO houses IPMs extracted from published peer-reviewed studies, as well as extensive metadata that enables researchers to filter the database down to models that fit their research questions. `Rpadrino` provides a clean interface to rebuild these models from _R_ using the `ipmr` _R_ package as a backend. This chapter also includes two detailed case studies on how to use these tools for both standalone analyses and in conjunction with other databases to power much larger scale syntheses.

In **Chapter 4**, I examine the drivers of invasiveness in the _Carpobrotus_ genus using data collected by unmanned aerial vehicles (UAVs, drones) on four continents. We show how climate impacts the population dynamics of this widely invasive genus using a combination of IPMs and Life Table Response Experiments (LTREs). We further show that the optimal climate conditions for this species are unlikely to be realized at any single site, as the drivers examined display differing response curves which do not share optima (WILL NEED REVISION BUT THIS IS MY FIRST IMPRESSION). This chapter further demonstrates how to scale inference across levels of organization (individuals to populations) and space (local to regional, regional to continental). 
 
# Chapter 1: Introduction

### Population models in ecology

Understanding population dynamics is central to ecology (Crone et al. 2011), evolution (Metcalf & Pavard 2007), and conservation (Morris & Doak 2002). Survival and reproduction rates form the two main fitness components that drive evolution, and determine the fate of populations as a whole (Coulson 2012). These vital rates are therefore crucial to understanding numerous other phenomena, such as change in population size, changes in causes of mortality and/or fertility, or the direction and magnitude of natural selection. Indeed, the utility of this idea - that the fates of closed populations are determined by rates of birth and death - has been demonstrated numerous times over the years. Notable applications include (though are certainly not limited to) understanding predator-prey interactions (Lotka 1925, Volterra 1926), the consequences of age/stage/size structure in populations (Leslie 1945, Lefkovitch 1965), invasive and endangered species management (Caswell 1978, de Kroon et al. 1986, Morris & Doak 2002),  and the consequences of climate change (Teller et al. 2016, Compagnoni et al. 2021). 

Early applications of population biology focused on human populations and modeled total population size (rather than, for example, number of age $a$ individuals). These yielded insight into drivers of exponential population growth (and an unhealthy dose of nativist and racist sentiment, Franklin 1751, Malthus 1798). They gave the classic equation: 

$$N(t) = N_0e^{rt}. \tag{1.1.1}$$

$N(t)$ is the population size at time $t$, $N_0$ is the initial population size, and $r$ is the intrinsic rate of population growth. Subsequent developments of logistic growth to account for carrying capacity were introduced by Verhulst in 1838, and (perhaps more famously) applied by A.G. McKendrick (1911) to study bacterial growth: 

$$N(t+1) = (1 + r (1 - \frac{N(t)}{K}))N(t). \tag{1.1.2}$$

Here, $K$ represents the carrying capacity of the population, which is the maximum population size that the environment can support. Shortly afterwards, biologists began to recognize the importance of dividing the population into sub-groups to understand how within population differences in vital rates impacts dynamics. In other words, not all individuals in the $N(t)$ terms are identical. The practice of sub-dividing the population based on some trait is refered to as "structuring the population". Beginning in the early 1900s, theoretical advancements aimed to understand age-specific rates of birth and death is recognized as the beginning of this field of study (McKendrick 1926):

$$\frac{\delta n}{\delta t} + \frac{\delta n}{\delta a} = -m(a) n \tag{1.1.3}$$

Here, $m(a)$ is the age specific mortality rate. However, this approach did not become popular in ecology for quite some time, likely due to the tedious nature of the calculations (Ebert 1999). P.H. Leslie introduced the discrete time, age structured projection matrix in 1945 (Leslie 1945). This approach provided a more elegant way of summarizing vital rates, though still required dutiful attention to detail when performing the calculations by hand: 

$$N(a, t+1) = AN(a,t). \tag{1.1.4}$$

$N(a, t)$ and $N(a, t+1)$ represent the population age distribution at time $t$ and $t+1$, and $A$ is an $m \times m$ matrix with $m$ age classes and their associated age-specific survival and fertility rates (where $m$ is maximum age $a$ of the organism in question). Extensions of Leslie's age-based approach to incorporate stage-structured populations (Lefkovitch 1965) and multiple structuring variables (Goodman 1969) advanced theory, but did not necessarily drive broader uptake. The advent of personal computers and the recognition that numerical analysis methods fit well with this rapidly growing technology lead to a rapid adoption in the late 1970s and early 1980s. Concurrently, Hal Caswell and colleagues' showed how to calculate population responses to absolute (sensitivity, 1978) and proportional (elasticity, 1986) perturbations to matrix elements, which are still a vital analytical tool for applied practitioners and theoreticians. A subsequent text book provided computer code to implement these analyses, along with many others, which helped make these approaches more accessible (Caswell 1989).

In the intervening years, further developments have enabled researchers to accommodate additional biological realism into their structured population models, such as density dependence (Leslie 1959, Liu & Cohen 1987, Jensen 1995) and environmental stochasticity (Caswell 1989). However, there was also a recognition that many species have their life cycle structured by a continuous trait, like weight, height, diameter, or hatching date. Furthermore, partitioning these continuous traits up into discrete categories introduced increased data requirements for parameterization and/or reduced the biological realism of the model by treating all individuals in a given class as identical (Vandermeer 1978). The integral projection model (IPM) was proposed to alleviate this issue (Easterling et al. 2000). IPMs provide a framework to model the dynamics of populations structured by any number of continuous and discrete traits (Ellner & Rees 2006, Ellner et al. 2016). Further developments to the IPM framework yielded methods for including a variety of additional factors, including density dependence (Rose et al. 2005, Adler et al. 2010), environmental stochasticity (Childs et al. 2004, Rees & Ellner 2009), spatial structure (Jongejans et al. 2011), and demographic stochasticity (Vindenes et al. 2011). These developments have enabled analyses across a variety of fields, including evolution (e.g. Coulson et al. 2011), conservation biology (e.g. Ferrer-Cervantes et al. 2012), and global change biology (e.g. Nicole et al. 2011, Simmonds et al. 2020). 

The main strengths of IPMs lie in i) the ability to choose continuous traits to model populations, ii) the ability to flexibly incorporate discrete and continuous traits into the same model, iii) efficient parameterization of the model using regression-based toolkits, and iv) numerical computation of the integrals (see $1.2$) mean tools available for matrix models are usually available for IPMs. Thus, there is a rich array of life cycles one can model, with a variety of well understood tools for fitting functional forms to real data, and a broad repertoire of tools and frameworks to analyze the resulting IPM with.

### Simple Integral Projection Models

An IPM describes how the abundance and distribution of a _trait_ (denoted $z$ and $z'$) changes from time $t$ to time $t+1$. Simple IPMs are IPMs that use one, and only one, trait to structure the population in question. The current trait distribution is given by the function $n(z,t)$. A simple IPM for the trait distribution of $z'$ at $t+1$ is then: 

$$n(z', t + 1) = \int_L^U K(z',z)n(z, t)dz. \tag{1.2.1}$$

$K(z',z)$ is a _kernel_ function (often referred to as a kernel) that describes all possible transitions of existing individuals (survival and maturation) and recruitment of new individuals ((a)sexual reproduction) from $t$ to $t+1$. $L,U$ are the lower and upper bounds that the value of trait $z$ can take, and the interval $[L, U]$ is referred to as the _domain_. To compute total population size, we simply integrate the trait distribution functions $\int_L^Un(z,t)$ ($\int_L^U n(z',t+1)$ to get the population size at $t+1$).

To aid interpretability, the kernel $K(z',z)$ is usually decomposed into _sub-kernels_ representing transitions related to existing individuals (i.e. survival, change in trait value) and creation of new individuals:

$$K(z',z) = P(z', z) + F(z', z) + C(z',z). \tag{1.2.2}$$

$P(z',z)$ describes transitions of existing individuals due to survival and change in trait value. $F(z',z)$ describes per-capita sexual reproduction contributions. $C(z',z)$ describes per-capita asexual reproduction contributions. IPM theory assumes these sub-kernels are somewhat smooth functions, with piecewise continuity satisfying this assumption (Ellner et al. 2016, Ch 2.2).

The $P(z',z)$, $F(z',z)$, and $C(z',z)$ sub-kernels are in turn comprised of vital rate functions. It is at this step that IPMs link the researcher's data to the population dynamics. For most IPMs, these vital rate functions are generated via regression models fit to real data collected in the field or in the lab (or some combination thereof). 

For example, the $P(z',z)$ kernel from Bogdan et al. (2021) for _Carpobrotus spp._ contains two vital rate functions: 

$$P(z',z) = s(z) G(z',z). \tag{1.2.3}$$

$s(z)$ was parameterized using a logistic regression of survival outcome (either 0 (dead) or 1 (alive)) at $t+1$ on plant size $z$. $G(z',z)$ was parameterized using a linear regression of size $z'$ on size $z$. This resulted in:

$$Logit(s(z)) = \beta_{0,s} + \beta_{1,s} * z, \tag{1.2.4}$$

$$G(z',z) = f_G(z' | \mu_G(z), \sigma_G), \tag{1.2.5}$$

$$\mu_G(z) = \beta_{0,G} + \beta_{1,G} * z. \tag{1.2.6}$$

$\beta_{0,i}$ and $\beta_{1,i}$ are intercepts and slopes from each regression, respectively, and  $f_G$ in $1.2.6$ denotes a Gaussian probability density function. $\sigma_G$ is the standard deviation of the residuals from the growth regression. 

The integrals are usually impossible to solve analytically (Ellner & Rees 2006), so numerical approximations are computed instead. The midpoint rule is the most commonly used one (others are possible, see Ellner et al. 2016, Chapter 6). This divides the domain $[L,U]$ into $m$ artificial bins centered at $z_i$ with a width of $h = (U - L)/m$, and midpoints $z_i = L + (i - 0.5)*h$ for $i = 1, 2, \dots, m$. The midpoint rule approximation of $1.2.1$ becomes:

$$n(z_j, t+1) = h \sum\limits_{i=1}^m K(z_j, z_i)n(z_i,t). \tag{1.2.7}$$

The numerical integration creates a discretized projection matrix $K$ analogous to $A$ in $1.1.4$. Matrix multiplication of the discretized kernel and the discretized trait distribution generates a new trait distribution. 


```{r echo = FALSE, warning = FALSE, message = FALSE, fig.height = 8, fig.width = 8, fig.cap = 'Figure 1: The link between data (A-B) and an IPM (C-D). This is typically a reversal of the way IPMs are described above, in that researchers start with data on sizes and demographic outcomes (A). These data get turned into functions in 1.2.4 - 1.2.6 using a variety of statistical methods, most typically a regression (B). In B, the solid black line depicts the predicted mean size at _t+1_ vs observed size at _t_. The red lines depict the variation around the predicted mean. Note that the variance shrinks as the initial size increases - this illustrates the power of IPMs to accommodate a variety of functional relationships. Once individual vital rate functions are computed, these are combined to form P, F, and K (C) found in equations 1.2.2 and 1.2.3. From there, some biological inference is conducted (D). In this case, the per-capita growth rate ($\\lambda$) and its confidence interval are depicted.'}

library(ipmr)
library(nlme)

data(iceplant_ex)

ip_dat <- subset(iceplant_ex, log_size_next > -5 | is.na(log_size_next))

z  <- c(ip_dat$log_size, ip_dat$log_size_next)
zz <- seq(-4.5, 
          4,
          length.out = 100)

gr_mod <- gls(log_size_next ~ log_size, 
              data = ip_dat,
              weights = varExp(),
              na.action = na.omit)
mu_pr <- coef(gr_mod)[1] + coef(gr_mod)[2] * zz

# This first one has the wrong form, but makes for a pretty graph
sd_1  <- sqrt(exp(gr_mod$modelStruct$varStruct * zz[15])) - 0.2
sd_2  <- sqrt(exp(2 * gr_mod$modelStruct$varStruct * zz[50]))
sd_3  <- sqrt(exp(2 * gr_mod$modelStruct$varStruct * zz[85]))

y_1 <- dnorm(zz, mu_pr[15], sd_1)
y_2 <- dnorm(zz, mu_pr[50], sd_2)
y_3 <- dnorm(zz, mu_pr[85], sd_3)

s_mod <- glm(survival ~ log_size, data = ip_dat, family = "binomial")
p_mod <- glm(repro ~ log_size, data = ip_dat, family = "binomial")
f_mod <- glm(flower_n ~ log_size, data = ip_dat, family = "poisson")

g_pars <- setNames(as.list(coef(gr_mod)), paste0("g_", c("int", "slope"))) %>%
  c(list(g_v_slope = gr_mod$modelStruct$varStruct[1]))
s_pars <- setNames(as.list(coef(s_mod)), paste0("s_", c("int", "slope")))
p_pars <- setNames(as.list(coef(p_mod)), paste0("p_", c("int", "slope")))
f_pars <- setNames(as.list(coef(f_mod)), paste0("f_", c("int", "slope")))

rc_list <- list(rc_mu = mean(ip_dat$log_size_next[is.na(ip_dat$log_size)]),
                rc_sd = 0.75)

data_list <- c(g_pars, s_pars, p_pars, f_pars, rc_list)

ipm <- init_ipm("simple", "di", "det") %>%
  define_kernel(
    name = "P",
    formula = s * g,
    family = "CC",
    s = 1 / (1 + exp(-(s_int + s_slope * z_1))),
    g = dnorm(z_2, g_mu, g_sd),
    g_sd = sqrt(exp(2 * g_v_slope * z_1)),
    g_mu = g_int + g_slope * z_1,
    data_list = data_list,
    states = list(c("z")),
    evict_cor = TRUE,
    evict_fun = truncated_distributions("norm", "g")
  ) %>%
  define_kernel(
    name = "F",
    formula = p_rep * f * r,
    family = "CC",
    p_rep = 1 / (1 + exp(-(p_int + p_slope * z_1))),
    f = pmin(0.7, exp(f_int + f_slope * z_1)),
    r = dnorm(z_2, rc_mu, rc_sd),
    data_list = data_list,
    states = list(c("z")),
    evict_cor = TRUE,
    evict_fun = truncated_distributions("norm", "r")
  ) %>%
  define_impl(
    make_impl_args_list(
      kernel_names = c("P", "F"),
      int_rule = rep("midpoint",2),
      state_start = rep("z", 2),
      state_end =  rep("z", 2)
    )
  ) %>%
  define_domains(z = c(-4.5, 4, 100)) %>%
  define_pop_state(n_z = rep(1/100, 100)) %>%
  make_ipm()

k <- ipm$sub_kernels$P^1.2 + ipm$sub_kernels$F



par(mfrow = c(2, 2))
plot(log_size_next ~ log_size, 
     data = ip_dat, 
     xlim = c(-4.5, 4), ylim = c(-4.5, 4),
     xlab = "Size (t)",
     ylab = "Size (t + 1)")
text(x = -4.3, y = 3.4, labels = "A", cex = 1.3)

plot(log_size_next ~ log_size, 
     data = ip_dat, 
     xlim = c(-4.5, 4), ylim = c(-4.5, 4),
     xlab = "Size (t)",
     ylab = "Size (t + 1)",
     col = "grey60")
lines(zz, mu_pr, col = "black")
lines(zz[15] + y_1, zz, lwd = 2.2, lty = 2, col = "red")
lines(zz[50] + y_2, zz, lwd = 2.2, lty = 2, col = "red")
lines(zz[85] + y_3, zz, lwd = 2.2, lty = 2, col = "red")
text(x = -4.3, y = 3.4, labels = "B", cex = 1.3)


labs <- round(c(zz[1], zz[21], zz[41], zz[61], zz[81], zz[100]), 2)

image(t(k), xaxt = "n", yaxt = "n",
      ylab = "Size (t + 1)",
      xlab = "Size (t)")
axis(side = 1, at = seq(0, 1, by = 0.2), labels = labs)
axis(side = 2, at = seq(0, 1, by = 0.2), labels = labs)
text(x = 0.05, y = 0.925, labels = "C", cex = 1.3)

lams <- seq(0.95, 1.02, length.out = 50)
lam  <- 0.98

plot(1, lam, pch = 19, xlab = "", 
     ylab = "Population growth rate", ylim = c(0.93, 1.04), cex = 1.5,
     xaxt = "n")
axis(1, labels = FALSE, tick = FALSE)
lines(rep(1, 50), lams, lty = 3, lwd = 3)
abline(h = 1, col = "grey40", lty = 2, lwd = 2)
text(x = 0.625, y = 1.035, label = "D", cex = 1.3)


```

Because of the smiliarity between $K(z_j, z_i)$ in $1.2.7$ and $A$ in $1.1.4$, many of the tools developed for matrix models can be applied to IPMs with little to no additional effort. These include (but are not limited to!) population growth rates ($\lambda$), stable trait distributions ($w$), reproductive values ($v$), sensitivity and elasticity analysis. Statistical and mathematical methods for incorporating environmental stochasticity (Rees & Ellner 2009) and density dependence (Rose et al. 2005, Adler et al. 2010) are available. Furthermore, there are guides available for assistance with the former (Metcalf et al. 2015) and the latter (Ellner et al. 2016, Ch 5).

Simple IPMs have been developed for a wide array of applications, ranging from the theoretical to the practical. For example, Metcalf and colleagues (2009) combined simple IPMs for six tropical tree species with a discretely varying light environment to model life expectancy, age specific mortality, and passage times to a given size conditional on initial state. Bassar and colleagues (2016) developed a simple, density dependent IPM to show how asymmetric competition affects various life histroy traits of _Poecilia reticulata_). Nicole and colleagues (2011) examined how climate change and local habitat conditions would influence the viability of the endangered _Dracocephalum austriacum_. I developed a simple, density-independent IPM for _Lonicera maackii_ to show that the degree of competitive release was a function of phylogenetic and functional novelty in its new local community (Levin et al. 2019, Levin et al. 2020). Zucaratto and colleagues (2020) developed a simple IPM for the invasive _Roystonea oleracea_ to develop a management plan for preserving Ilha Grande, an island off the coast of Brazil whose Atlantic Forest habitat is threatened. Despite the wide application of simple IPMs, there is a need to model life cycles and species that aren't readily summarized by a single, continuous trait (e.g. plants with a discrete seed bank stage). To that end, Ellner & Rees (2006) developed the general IPM, which is covered in the next section.

### General Integral Projection Models

General IPMs are an extension of simple IPMs that allow researchers to model a population structured by multiple continuously and/or discretely traits. For example, many plants have long-lived seed banks, and researchers are typically consider this a discrete stage because of low trait variation across individuals, or because they lack sufficient data on said trait variation to create a continuous trait distribution (but see Eager et al. 2013). Therefore, the seedbank is a discrete stage represented by a single number (the number of seeds that are in it). Similarly, age- and trait-structured models may contain many copies of the same trait distribution - one for each age class. For example, an age $\times$ size structured IPM for _Ovis aries_ may have 21 copies for log transformed body weight denoted $z_a$ for $a \in [0,20]$ (Ellner, Childs, & Rees 2016, Ch 6). General IPMs allow researchers to accommodate these more complex life cycles in a single model.

The trait space in a general IPM is no longer one dimensional. We represent the entirety of the space with $Z$. $Z$ can include discrete points $D = \{z_1, z_2, \dots, z_D\}$, and a set of continuous domains $C = \{Z_{D+1}, Z_{D+2}, \dots, Z_{D+C}\}$ (Ellner et al. 2016, Chapter 6). Each continuous domain in $C$ is either a closed interval $[L,U]$ (like in a simple IPM), or a closed finite rectangle (e.g. with individuals cross-classified by size $z$ and "individual quality" $q$). Similarly, the population state is now comprised of different components. Discrete states are represented $n_j(t), j = 1, 2, \dots D$ which give the number of individuals with state $D$. Continuous components are represented $n_j(z_j, t), j = D + 1, D + 2, \dots D + C$, where the integral $\int_{Z_j} n_j(z_j,t)d_{z_j}$ gives the number of  individuals with state $j$. Transitions within and between traits are defined by a set of sub-kernels $K_{ij}, 1 \leq i,j \leq D+C$. When individuals in with trait $Z_j(t)$ contribute to $Z_i(t+1)$, $K_{ij} \neq 0$. Thus, there are four possible kinds of kernel components:

1. Discrete to discrete transitions: $K_{ij}$ is a single number. This includes, for example, the fraction of seeds in a seedbank that do not germinate and survives to remain in the seedbank the following year, or, when insect eggs remain in stasis across time steps, neither hatching nor dying. 

2. Discrete to continuous transitions: $K_{ij} = k_{ij}(z')$. $k_{ij}(z_i')$ is a state distribution that gives the probability of trait value $z'$ at $t+1$, and is the same for all individuals in $Z_j$. This includes, for example, the size distribution of seedlings emerging from the seedbank, or the size of newly hatched insects. 

3. Continuous to discrete transitions: $K_{ij} = k_{ij}(z)$. $k_{ij}(z_j)$ is the state dependent per-capita contribution to state $i$ at $t+1$ from state $j$ at time $t$. This includes, for example, the number of seeds that a size $z_j$ plant produces that do not immediately germinate and enter the seedbank, or the number of insect eggs that a size $z_j$ adult produces that do not hatch immediately. 

4. Continuous to continuous transitions: $K_{ij} = K_{ij}(z_i', z_j)$. These are bivariate kernels functions like the ones introduced above in the simple IPM section. These include, for example, survival/growth and fecundity kernels of existing plants, or survival/growth kernels of adult insects structured by body length. It is worth noting that $i$ and $j$ don't need to be the same trait here. For example, trees may have a continuously structured seedling stage where seedlings are structured by height, and a continuously structured adult stage where adults are structured by diameter at breast height (DBH). In this case, seedlings may start a transition with height $z_j$ at $t$ and become adults with DBH $z_i'$ at $t+1$.

There are two assumptions made by the general model: i) there are a finite number of domains, and ii) all of the continuous domains are bounded. The first assumption is not overly restrictive, because states like ontogeny only have a few possible values. Age is bounded by the maximum lifespan of the species, which need not depend on rate of senescence (which some species do escape) but on the maximum observed age in the data used to parameterize the model (incorporating the biological reality that everything dies eventually). The second assumption is not restrictive unless the state in question is spatial location (in which case, see Ellner et al. 2016, Ch 8). Most species have a maximum size they can grow to (or, at least, a maximum size they regularly attain before something else kills them), and so imposing a closed domain on the model will not create unrealistic populations (Ellner, Childs & Rees 2016, Ch 6.3).  

General IPMs unlock an even broader range of applications, some going beyond traditional ecological questions. For example, IPMs have been embedded into the SIR epidemiological model to estimate onward transmission probability of malaria parasites based on within-host parasite load dynamics (Metcalf et al. 2015). Stepping back into the realm of more traditional ecology, Bruno et al. (2011) used a general IPM to model population dynamics of corals infected with a fungal pathogen. Invasion biology has also benefited from general IPMs, with a multitude of publications addressing mechanisms underlying their rapid population growth (e.g. Crandall & Knight 2017, Levin et al. 2020), rate of spread (e.g. Jongejans et al. 2011), probability of invasion given future climate (e.g. Merow et al. 2017), and potential control pathways (e.g. Erickson et al. 2017, Lommen et al. 2018). General IPMs have proven useful for quantifying how biotic (e.g. Adler et al. 2010, Simmonds et al. 2020) and abiotic interactions (Rees & Ellner 2009, Compagnoni et al. 2021) influence population dynamics. This is far from an exhaustive list, but hopefully illustrates the range of possibilities along the theoretical to applied continuum.

### Survey of available tools for IPM implementation

In parallel with developments in IPM theory, researchers have produced tools to assist with their implementation. For example, there are numerous "how-to" guides covering implementation of environmental variation (Metcalf et al. 2015), a practical guide for conservation biologists (Merow et al. 2013), and at least two _R_ packages to assist with creating some types of IPMs (Metcalf et al. 2013, Shefferson et al. 2021). Additionally, a book published by Ellner, Childs & Rees (2016) contains in-depth examinations of the theory underlying various IPM applications, and provides free _R_ code demonstrating how to implement these with real data. 

Additionally, at least two _R_ packages exist to help with fitting and implementing IPMs. `IPMPack` assists with both vital rate regression modelling and IPM implementation (Metcalf et al. 2013). It can handle a variety of IPM types, including simple IPMs, a subset of general IPMs, and includes functionality for specifying stochastic variation and conducting a variety of analyses, including life table response experiments, computing life expectancies, passage times, population growth rates, and perturbation based analyses. Furthermore, it contains excellent diagnostic tools to help users understand how their model is performing, and whether it may contain common flaws (e.g. lots of eviction, unrealistic size ranges, etc).  `lefko3` is an _R_ package offering a similar level of abstraction as `IPMPack` (i.e. handles both statistical modelling and IPM implementation). `lefko3` is designed specifically to help with analyses that incorporate an individual's history into the future state (i.e. $n(z, t +1) = f(n(z,t), n(z, t-1))$, Shefferson et al. 2021). It is more general than `IPMPack` in that it can handle a greater range of data non-independence (i.e. with mixed effects models) and a greater range of distributions for $z$. Both packages require a fixed set of vital rate functional forms, and rely on other _R_ packages, such as `nlme`, `lme4`, `VGAM`, and `glmmTMB`. This introduces a tradeoff between perceived ease of use on the one hand, and flexibility in modeling single vital rates (and by extension, the life cycle as a whole) on the other hand. This tradeoff represents a substantial gap in the available tooling at our disposal, because pressing research questions demand more flexibility in IPM parameterization and underlying functional forms. 

Much of this required complexity arises from the need to faithfully represent vital rates and their dependence not just on the trait (or traits) in the IPM, but external environmental factors as well. These external factors can include (but are not limited to) other species or changing climate conditions. Furthermore, the responses need not be linear (in fact, they often may not be). These issues are actually a subset of general regression modelling problems, and the field of computational statistics is constantly developing new tools to address them. For example, `mgcv` implements generalized additive models to handle estimation of semi-parametric smooth functions using maximum likelihood (Wood 2017). `lme4` similarly uses maximum likelihood to estimate linear mixed models to handle non-independence in the underlying data (e.g. from spatial clustering, phylogenetic structure, or nested experimental design, Bates et al. 2015). `brms` has become a popular interface to the `Stan` library for estimating Bayesian models with MCMC algorithms (Buerkner 2018, Stan Development Team 2020). `brms` is particularly popular because of it provides a familiar _R_ formula interface and allows users to specify nearly any type of model, linear or non-linear, that they desire. It also leverages `mgcv`'s basis-generating functions (and the duality of certain spline basises as random effects) to estimate GAMs in a fully Bayesian framework. 

There is an astounding array of additional tools researchers may use to estimate trait-demography relationships, and this is only a brief overview. However, population biologists do not have a framework that allows us to harness the full breadth of these tools in an IPM context without writing all analysis code from scratch. This is problematic because 1) untested IPM code may contain bugs, 2) it potentially creates barriers to access for early career researchers who may not have a strong support network to teach them how to create IPMs, and 3) it creates problems in performing syntheses because we have no common data structure to represent IPMs and their underlying functional forms.  

### Demographic data for broader syntheses

Given the age and maturity of modelling structured populations, researchers have become interested in compiling databases of published literature for synthesis applications. COMPADRE (Salguero-Gomez et al. 2014), COMADRE (Salguero-Gomez et al. 2015), and DatLIFE (DatLIFE 2018) are the most directly relevant efforts for this dissertation. Others, including popler (Compagnoni et al. 2019), BIEN (Maitner et al. 2017), and $\mathrm{D}^3$ (Hintze et al. 2013) are all important sources of information, but do not necessarily incorporate population structure or the entire life cycle of the species into their data. Still others, such as the HMD (Human Mortality Database 2022), HLD (Human Lifetable Database 2022), and IPUMS Terra (Ruggles et al. 2018) are important contributions, but limit their phylogenetic scope to _Homo sapiens_. 

COMPADRE and COMADRE are the most closely linked databases to this dissertation. Their history stretches back to the late 1980s, when Silvertown and Franco began archiving published MPMs as time- and space-averaged matrix models. The synthetic rewards included, for example, the concept of the fast-slow continuum of plant strategies (Franco & Silvertown 1997), a broad examination of the relative contributions of vital rates to population growth rates (Silvertown et al. 1993), and a synthesis of the evolution of senesence in plants (Silvertown et al. 2001). Researchers spent much of the 2000s working in parallel to compile similar data sets until unifying under the umbrella of COMPADRE (and expanding to mammals via COMADRE) in 2011 (Salguero-Gomez et al. 2014). In subsequent years, these databases have been enhanced with additional data and metadata to assist researchers in linking these data to other data sources to address increasingly complicated questions.  Since becoming open access in 2014 (COMPADRE) and 2015 (COMADRE), they have been cited 157 and 77 times, respectively. Clearly, these data sources have provided a substantial return on investment. Some of these returns include quantifiying the diversity of rates of ageing (Jones et al. 2014), plant and mammal species responses to climate change (Compagnoni et al. 2021, Paniw et al. 2021), and the expansion of the fast-slow continuum theory to include a reproductive strategy axis (Salguero-Gomez et al. 2015).

IPM usage is growing rapidly (Levin et al. 2021). IPMs are now the tool of choice to analyze populations that are structured by continuous traits, because the alleviate the need to construct arbitrary bins that discretize these continuous traits (e.g. create size classes for a tree species, Ellner & Rees 2006). Given the success of MPM repositories, a database of IPMs is necessary for synthetic studies in ecology. However, there are unique challenges posed in implementing such a database. Existing approaches to digitizing MPMs entail entering the iteration matrix/sub-matrix elements into databases (i.e. entering that actual value of the matrix cell). This is feasible for MPMs, sice most do not have many rows and columns (the majority of COMPADRE and COMADRE matrices are less than $20 \times 20$). Furthermore most transition elements are on the order of $[0.01, 1]$, making them easier to enter by hand. IPM iteration matrices are much higher dimensional (on the order of $50 \times 50 - 1,000 \times 1,000$), and the transition elements can be much smaller (i.e. $[1^{-10}, 1^{-2}]$). Entering the latter by hand would inevitably result in errors, and might not even be possible since researchers rarely publish raw projection kernels (SC Levin, personal observation). Finally, capturing the IPM's underlying functional form, rather than just the iteration matrices/sub-matrices, enables more applications. Thus, in addition to the usual metadata and parameter value digitization that comes with creating a database, there are two extra tasks that must be completed to have a working IPM database:
 
1. Define a syntax to represent IPMs symbolically in a database. This representation needs to be flexible enough to accommodate models across the range of complexities in the existing literature, and able to handle as much forthcoming, additional complexity as possible. 

2. Implement an engine that can transform this database representation into something that's actually useful for a researcher who is not interested in the underlying technical details, and simply wants to focus on their research questions (i.e. most researchers). Given the dominance of _R_ (R Core Development Team 2022) in the field of ecology, this would be the logical choice for such an engine. However, to facilitate including a broader array of backgrounds, bindings to other languages would be ideal. Thus, the syntax in Objective 1 should not be overly specific to _R_, and follow more general mathematical notation.
 
### IPMs and consideration of scale

Given the range of applications that IPMs, it is worth stepping back and considering the various scales they might function across and how they have been applied to these scales so far. When using the term scales, I am referring broadly to time, space, and phylogeny. Within each of these, there are differing ways in which they apply:

1. Temporal scale, with respect to IPMs, usually comes in two flavors - the time step of the model iteration (i.e. the amount of actual time between $t$ and $t+1$), and the time frame of the total projection (i.e. stochastic models versus transient dynamic models versus asymptotic dynamics in deterministic models). In the former, the value is usually determined either by the speed in which the organism completes its life cycle, or, by the time step that has a meaningful biological interpretation. The latter is largely determined by the research question. Stochastic and asymptotic dynamics have dominated the literature over the years, but there is growing appreciation of the importance of transient dynamics.

2. Spatial scale, with respect to IPMs, typically refers to either the extent of a population, or the number of populations studied and how far apart they are, or both. The former can very greatly across study species. For example, the sampling grain for a population of _Canis lupus_ (Yellowstone National Park, Coulson et al. 2011) can very different from the sampling grain for _Carduus nutans_ ($1m \times 1m$ plots, Levin et al. 2019). For the latter, there are numerous considerations, largely driven by the research question. For example, consideration of competitive interactions across many sites would require separate models fo, as competitive pressure decays rapidly with distance between individuals. Consideration of weather drivers of population dynamics requires sampling a broad spatial extent at small spatial grains to capture variation in both weather and population performance. This is far from an exhaustive list, and is only meant to illustrate that space must be considered.  

3. Phylogenetic scale, with respect to IPMs, is not one that is often considered. However, it is a useful concept, as many questions in ecology and evolution seek to generalize from species to genus to family, etc. The vast, vast majority of IPMs quantify dynamics of a one species. Some may include feedbacks between species (e.g. Bruno et al. 2011, Adler et al. 2010). However, there is some work that examines "family-level" demography (Traill et al. 2021).

Invasive species, in addition to presenting a pressing ecological (XXXX) and economic threat (XXXX), present an ideal system to examine all three of these meanings of scale. Species that become invasive usually undergo rapid spread, presenting a series of transient dynamics at the leading edge, while converging to asymptotic dynamics in the older center of invasion (XXX). The spatial scale of invasions can be viewed as local, in which the individuals in the population come to dominate their immediate area, and large extents (e.g. regional, continental) into which they spread and establish larger populations (XXX). Phylogenetic scale is especially prominent in the context of invasion biology, with work spanning nearly every taxonomic level of organization from single species (XXX), collections of species (e.g. XXXX), genera (e.g. XXXX), families (XXX), and even the entire angiosperm clade (Darwin 1859). Despite all of this, there is still much debate over how these various meanings of scale should apply to invasion biology.

### Objectives of the dissertation

The first objective of this dissertation is to provide a consistent interface for implementing IPMs. `ipmr` (**Chapter 2**) provides an engine for implementing IPMs from symbolic representations. It imposes no restrictions on vital rate models - researchers may use the full breadth of _R_'s modelling capabilities. Furthermore, it provides tools for basic analyses, and is extensively documented with examples of more complicated ones. The latter point substantially reduces difficulty in beginning to use the package, and helps researchers focus on their research questions rather than details of programming. Chapter two is published as: Levin, S.C., Childs, D.Z., Compagnoni, A., Evers, S., Knight, T.M. & Salguero-Gomez, R (2021) ipmr: flexible implementation of Integral Projection Models in R. Methods in Ecology and Evolution 12(10): 1826-1834.

The second objective of this dissertation is to provide a database of peer-reviewed published IPMs that researchers may use for large scale synthesis. This breaks down into a two sub-objectives: 

1. Provide a central repository with consistent representations of peer-reviewed published IPMs. This is provided via PADRINO (**Chapter 3**).

2. Provide a tool that abstracts away the details of the database and enables researchers to easily subset and reproduce these IPMs. This is provided via `Rpadrino` (**Chapter 3**) and `ipmr` (**Chapter 2**). 

PADRINO provides a repository for IPMs with extensive metadata. It also contains symbolic representations of the IPMs and the parameters used to implement them. It can store determinstic or stochastic, density/frequency-dependent or -independent, and simple or general IPMs. The syntax used to store the IPMs symbolically mirrors the mathematical notation for them. `Rpadrino` provides an interface to download PADRINO, subset PADRINO, and translate PADRINO's syntax into valid `ipmr` code. `ipmr` is then used as backend for reconstructing IPM objects. Chapter 3 is published as: Levin, S.C., Evers, S., Potter, T., Guerrero, M.P., Childs, D.Z., Compagnoni, A., Knight, T.M. & Salguero-Gomez, R. (2022) Rpadrino: an R package to access and use PADRINO, an open access database of Integral Projection Models. Methods in Ecology and Evolution. DOI: https://doi.org/10.1111/2041-210X.13910. 

**Chapter 4** of this dissertation applies these tools to examine how environmental drivers affect demography of invasive species at both local and broad spatial scales using the _Carpobrotus_ genus. I used drones to collect demographic data from 14 sites on four continents across the native and invaded range. I develop an IPM using `ipmr` for each site that includes environmental covariates to show how _Carpobrotus_ responds to environmental drivers. Finally, I use a Life Table Response Experiment to quantify the contributions of climate drivers to overall fitness.

In **Chapter 5**, I provide a synthesis of findings and recommendations for future research.

### Citations

1. Adler, P.B., Ellner, S.P. & Levine, J.M. (2010). Coexistance of perennial plants: an embarassment of niches. Ecology Letters 13: 1019-1029. https://doi.org/10.1111/j.1461-0248.2010.01496.x

2. Bassar, R.D., Childs, D.Z., Rees, M., Tuljapurkar, S., Reznick, D.N. & Coulson, T. (2016) The effects of asymmetric competition on the life history of Trinidadian guppies. Ecology Letters 19(3): 268-278. 

4. Bogdan, A., Levin, S.C., Salguero-Gómez, R., Knight, T.M. (2021). Demographic analysis of Israeli Carpobrotus populations: management strategies and future directions. PLoS ONE 16(4): e0250879.  https://doi.org/10.1101/2020.12.08.415174

1. Caswell, H. (1978). A general formula for the sensitivity of population growth rate to changes in life history parameters. Theoretical Population Biology 14(2): 215-230. https://www.jstor.org/stable/2528566

2. Caswell, H. (1989). Matrix Models 1st edition.

3. Childs, D.Z., Rees, M., Rose, K.E., Grubb, P.J., & Ellner, S.P. (2004). Evolution of size-dependent flowering in a variable environment: construction and analysis of a stochastic integral projection model. Proceedings of the Royal Society B 271(1547): 425-434. https://doi.org/10.1098/rpsb.2003.2597

8. Compagnoni, A., Levin, S.C., Childs, D.Z., Harpole, S., Paniw, M., Roemer, G., Burns, J.H., Che-Castaldo, J., Rueger, N., Kunstler, G., Bennett, J.M., Archer, C.R., Jones, O.R., Salguero-Gómez, R., & Knight, T.M. (2021). Herbaceous perennial plants with short generation time have stronger responses to climate anomalies than those with longer generation time. Nature Communications 12: 1824. https://doi.org/10.1038/s41467-021-21977-9

9. Compagnoni, A., Bibian, A.J., Ochocki, B.M., Rogers, H.S., Schultz, E.L., Sneck, M.E., Elderd, B.D., Inouye, D.W., Jacquemyn, H. & Miller, T.E.X. (2016) The effect of demographic coorelations on the stochastic population dynamics of perennial plants. Ecological Monographs 86(4): 480-494.

10. Compagnoni, A., Bibian, A.J., Ochocki, B.M., Levin, S.C., Zhu, K. & Miller, T.E.X. (2019) popler: an R package for extraction and synthesis of population time series from long-term ecological research (LTER) network. Methods in Ecology and Evolution 11(2): 258-264.

23. Coulson, T., MacNulty, D.R., Stahler, D.R., von Holdt, B., Wayne, R.K., & Smith, D.W. (2011). Modeling effects of environmental change on wolf population dynamics, trait evolution, and life history. Science 334(6060): 1275-1278. https://doi.org/10.1126/science.1209441

12. Crandall, R.M. & Knight, T.M. (2017). Role of multiple invasion mechanisms and their interaction in regulating the population dynamics of an exotic tree. Journal of Applied Ecology 55(2):885-894. https://doi.org/10.1111/1365-2664.13020

20. Crone, E.E., Menges, E.S., Ellis, M.M., Bell, T., Bierzychudek, P, Ehrlen, J. et al. (2011) How do ecologists use matrix population models? Ecology Letters 14(1): 1-8. DOI: https://doi.org/10.1111/j.1461-0248.2010.01540.x

22. Darwin, C. (1859) On the Origin of Species By Means of Natural of Selection. John Murray: UK.

21. Doak, D. & Morris W.F. (2002). Quantitative Conservation Biology: Theory and Practice of Population Viability Analysis. Oxford, UK: Oxford University Press

5. Easterling, M.R., Ellner, S.P., & Dixon, P.M. (2000). Size-specific sensitivity: Applying a new structured population model. Ecology 81(3): 694-708. https://doi.org/10.1890/0012-9658(2000)081[0694:SSSAAN]2.0.CO;2

2. Ebert, T.A. (1999). Plant and Animal Populations: Methods in Demography. Academic Press, San Diego, California.

16. Ellner, S.P. & Rees, M. (2006). Integral Projection Models for species with complex demography. The American Naturalist 167(3): 410-428.

7. Ellner, S.P., Childs, D.Z., & Rees, M. (2016) Data driven modelling of structured populations: a practical guide to the Integral Projection Model. Springer, Switzerland.

12. Erickson, R.A., Eager, E.A., Brey, M.B., Hansen, M.J., & Kocovsky, P.M. (2017). An integral projection model with YY-males and application to evaluating grass carp control. Ecological Modelling 361: 14-25. https://doi.org/10.1016/j.ecolmodel.2017.07.030

55. Franco, M. & Silvertown, J. (1997) Life History Variation in Plants: An Exploration of the Fast-Slow Continuum Hypothesis. Cambridge University Press, Cambridge, UK, New York, NY, USA.

19. Franklin, B. (1751) Observations concering the increase of mankind, peopling of countries, etc. 

47. Ferrer-Cervantes, M.E., Mendez-Gonzalez, M.E., Quintana-Ascencio, P-F., Dorantes, A., Dzib, G., & Duran, R. (2012). Population dynamics of the cactus _Mammillaria gaumeri_: an integral projection model approach. Population Ecology 54: 321-334. DOI: https://doi.org/10.1007/s10144-012-0308-7 

46. Goodman, L.A. (1969) The analysis of population growth when the birth and death rates depend upon several factors. Biometrics 25(4): 659-681.

10. Jensen, A.L. (1995) Simple density-dependent matrix for population projection. Ecological modelling 77: 43-48.

55. Jones, O.R., Scheuerlein, A., Salguero-Gomez, R., Camarda, C.G., Schaible, R., Casper, B.P., _et al._ (2014). Diversity of ageing across the tree of life. Nature 505: 169-173.

14. Jongejans, E., Shea, K., Skarpaas, O., Kelly, D., & Ellner, S.P. (2011). Importance of individual and environmental variation for invasive species spread: a spatial integral projection model. Ecology 92(1): 86-97. https://doi.org/10.1890/09-2226.1

4. de Kroon, H., Plaisier A., van Groenendael, J., & Caswell, H. (1986). Elasticity: the relative contribution of demographic parameters to population growth rate. Ecology 67(5): 1427-1431. https://doi.org/10.2307/1938700.

26. Human LifetableDatabase. University of California, Berkeley (USA), and Max Planck Institute for Demographic Research (Germany). Available at www.lifetable.de.

25. Human Mortality Database.  University of California, Berkeley (USA), and Max Planck Institute for Demographic Research (Germany). Available at www.mortality.org or www.humanmortality.de.

3. Lefkovitch, L.P. (1965). The study of population growth in organisms grouped by stages. Biometrics 21(1): 1-18. https://www.jstor.org/stable/2528348

5. Leslie, P.H. (1945) The use of matrices in certain population mathematics. Biometrika, 33(3): 183–212.

73. Levin, S.C., Crandall, R.M. & Knight, T.M. (2019) Population projection models for 14 alien plant species in the presence and absence of aboveground competition. Ecology 100(6): e02681. 

74. Levin, S.C., Crandall, R.M., Pokoski, T., Stein, C. & Knight, T.M. (2020) Phylogenetic and functional distinctiveness explain alien plant population responses to competition. Proceedings of the Royal Society B 287: 20201070.

75. Levin, S.C., Childs, D.Z., Compagnoni, A., Evers, S., Knight, T.M. & Salguero-Gomez, R (2021) ipmr: flexible implementation of Integral Projection Models in R. Methods in Ecology and Evolution 12(10): 1826-1834.

9. Liu, L. & Cohen, J.E. (1987). Equilibrium and local stability in a logistic matrix model for age-structured populations. Journal of Mathematical Biology 25: 73-88.

10. Lommen, S.T.E., Jongejans, E., Leitsch-Vitalos, M., Tokarska-Guzik, B., Zalai, M., Mueller-Schaerer, H. & Karrer, G. (2018) Time to cut: population models reveal how to mow invasive common ragweed cost-effectively. NeoBiota 39: 53-78. DOI: 10.3897/neobiota.39.23398

11. Lotka, A.J. (1925) Elements of Physical Biology, Williams and Wilkins.

12. Maitner B.S., Boyle, B., Casler, N., Condit, R., Donoghue, J., Duran, S.M. et al. (2017) The BIEN R package: a tool to access the Botanical Information and Ecology Network database. Methods in Ecology and Evolution 9(2): 373-379.

70. Hintze C., Heydel, F., Hoppe, C., Cunze, S., Koenig, A. & Tackenberg, O. (2013). D^3: The Dispersal and Diaspore Database - baseline data and statistics on seed dispersal. Perspectives in Plant Ecology, Evolution, and Systematics 15(3): 180-192.

20. Malthus, T. (1798) An essay on the principle of population. J. Johnson, London, England.

21. McKendrick, A.G. (1926) Applications of mathematics to medical problems. Proceedings of the Edinburgh Mathematical Society 44: 98-130

45. McKendrick, A.G. & Kesava Pai, M. (1911). The Rate of Multiplication of Micro-organisms: A Mathematical Study. Proceedings of the Royal Society of Edinburgh. 31: 649–653. doi: 10.1017/S0370164600025426.

22. Merow, C., Dahlgren, J.P., Metcalf, C.J.E., Childs, D.Z., Evans, M.E.K., Jongejans, E., Record, S., Rees, M., Salguero-Gomez R., & McMahon, S.M. (2014). Advancing population ecology with integral projection models: a practical guide. Methods in Ecology and Evolution 5: 99-110. https://doi.org/10.1111/2041-210X.12146S

27. Merow, C., Bois, S.T., Allen, J.M., Xie, Y. & Silander Jr., J.A. (2017) Climate change both facilitates and inhibits invasive plant ranges in New England. Proceedings of the National Academy of Sciences 114(16): E3276-E3284.

26. Metcalf et al. (2009) A time to grow and a time to die:

23. Metcalf, C.J.E., Ellner, S.P., Childs, D.Z., Salguero-Gómez, R., Merow, C., McMahon, S.M., Jongejans, E., & Rees, M. (2015). Statistical modelling of annual variation for inference on stochastic population dynamics using Integral Projection Models. Methods in Ecology and Evolution 6(9): 1007-1017. https://doi.org/10.1111/2041-210X.12405

65. Metcalf, C.J.E., Graham, A.L., Martinez-Bakker, M. & Childs, D.Z. (2015) Opportunities and challenges of Integral Projection Models for modelling host-parasite dynamics. Journal of Animal Ecology. DOI:  10.1111/1365-2656.12456

24. Metcalf, C.J.E., McMahon, S. M., Salguero-Gómez, R. & Jongejans, E. (2013). IPMpack: an R  package for integral projection models. Methods in Ecology and Evolution. 4(2): 195-200. https://doi.org/10.1111/2041-210x.12001

26. Metcalf, C.J.E. & Pavard, S. (2007) Why evolutionary biologists should be demographers. Trends in Ecology and Evolution 22(4): 205-212. DOI: 10.1016/j.tree.2006.12.001.

17. Nicole, F., Dahlgren, J.P., Vivat, A., Till-Bottraud, I. & Ehrlen, J. (2011). Interdependent effects of habitat quality and climate on population growth of an endangered plant. Journal of Ecology 99: 1211-1218.

13. Rees, M. & Ellner, S.P. (2009) Integral projection models for populations in temporally varying environments. Ecological Monographs 79(4): 575-594.

11. Rose, K.E., Louda, S.M. & Rees, M. (2005) Demographic and evolutionary impacts of native and invasive insect herbivores on _Cirsium canescens_. Ecology 86(2): 453-465.

71. R Core Team (2021). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.

26. Ruggles, S., Manson, S.M, Kugler, T.A., Haynes, D.A., Van Riper, D.C., Bakhtsiyarava, M. (2018). IPUMS Terra: Integrated data on population and environment: Version 2. Minneapolis, MN: IPUMS.

5. Salguero‐Gómez, R., Jones, O.R., Archer, C.R., Bein, C., de Buhr, H., Farack, C., Gottschalk, F., Hartmann, A., Henning, A., Hoppe, G., Roemer, G., Ruoff, T., Sommer, V., Wille, J. Voigt, J., Zeh, S., Vieregg, D., Buckley, Y.M., Che-Castaldo, J., Hodgson, D., et al. (2016) COMADRE: a global database of animal demography. Journal of Animal Ecology 85: 371-384. https://doi.org/10.1111/1365-2656.12482

6. Salguero-Gómez, R, Jones, O.R., Archer, C.A., Buckley, Y.M., Che-Castaldo, J., Caswell, C., Hodgson, D., Scheuerlein, A., Conde, D.A., Brinks, E., de Buhr, H., Farack, C., Gottschalk, F., Hartmann, A., Henning, A., Hoppe, G., Roemer, G., Runge, J., Ruoff, T., et al. (2014) The COMPADRE Plant Matrix Database: an online repository for plant population dynamics. Journal of Ecology 103: 202-218. https://doi.org/10.1111/1365-2745.12334

74. Salguero-Gómez, R., Jones, O.R., Jongejans, E., Blomberg, S.P., Hodgson, D.J., Mbeau-Ache, C., Zuidema, P., de Kroon, H., & Buckley, Y.M. (2015). Fast-slow continuum and reproductive strategies structure plant life history variation worldwide. Proceedings of the National Academy of Sciences 113(1): 230-235.

56. Silvertown, J., Franco, M. & Perez-Ishiwara, R. (2001) Evolution of senescence in iteroparous perennial plants. Evolutionary Ecology Research, 3, 393–412.

72. Silvertown, J., Franco, M., Pisanty, I. & Mendoza, A. (1993) Comparative plant demography: relative importance of life-cycle components to the finite rate of increase in woody and herbaceous perennials. Journal of Ecology, 81, 465–476.

29. Simmonds E.G., Cole, E.F., Sheldon, B.C., & Coulson, T. (2020). Phenological asynchrony: a ticking time-bomb for seemingly stable populations? Ecology Letters 23(12): 1766-1775. https://doi.org/10.1111/ele.13603

28. Shefferson, R.P., Kurokawa, S., & Ehrlen, J. (2020). LEFKO3: analysing individual history through size-classified matrix population models. Methods in Ecology and Evolution. https://doi.org/10.1111/2041-210X.13526

33. Teller, B.J., Adler, P.B., Edwards, C.B., Hooker, G. & Ellner, S.P. (2016) Linking demography with drivers: climate and competition. Methods in Ecology and Evolution 7: 171-183. DOI: 10.1111/2041-210X.12486

36. Traill, L., Plard, F., Galliard, J-M. & Coulson, T. (2021). Can we use a functional trait to construct a generalized model for ungulate populations? Ecology 102(4): e03289.

8. Vandermeer, J. (1978). Choosing category size in a stage projection matrix. Oecologia 32: 79-84.

9. Verhulst, P.H. (1838) Notice sur la loi que la population poursuit dans son accroissement. Corresp. mathématique et physique 10, 113–121

15. Vindenes, Y. Engen, S. & Saether, B-E. (2011) Integral projection models for finite populations in a stochastic environment. Ecology 92(5): 1146-1156. 

15. Volterra, V. (1926) “Variazioni e fluttuazioni del numero d’individui in specie animali conviventi”, Mem. Acad. Lincei Roma, 2, 31–113.

40. Zucarratto, R., Pires, A.S., Bergallo, H.G. & Portela, R.de C.Q. (2020) Felling the giants: integral projection models indicate adult management to control an exotic invasive palm. Plant Ecology. DOI: 10.1007/s11258-020-01090-5
