---
output:
  pdf_document: default
header-includes:
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
  - \usepackage{pdfpages}
---

\pagenumbering{gobble}

# Appendix 7: Supplementary Information for Chapter 4

# Model diagram

The following diagram describes a single iteration of the IPM.

\includepdf[pages = -, nup = 1, pagecommand = {}]{../Figures/time_step_diagram.pdf}

See the next section for a complete description of each vital rate parameter.

### IPM Equations

IPMs describe how the abundance and distribution of a continuously distributed trait changes in a population through discrete time. Vital rates are combined in projection kernels that describe state-dependent per-capita contributions of existing individuals to the population trait distribution in the following time step via survival and development (denoted ($P(z',z)$)) and sexual and asexual reproduction (denoted $F(z',z)$ and $C(z',z)$ respectively).


$$n(z',t+1) = \int_L^U [G(z'|z, \sigma, \theta) * s_a(z,\theta)]n(z,t)dz + s_s * r_a(z')sdl(t), \tag{4.1.1}$$

$$mf(t+1) = \int_L^U [p_f(z,\theta) * s_a(z,\theta) * r_f(z,\theta) * p_m * p_{fv}]n(z,t)dz, \tag{4.1.2}$$

$$sdl(t+1) = p_e * g_i * v_s * mf(t) + s_{sb} * g_{sb} * p_e * sb(t), \tag{4.1.3}$$

and 

$$sb(t+1) = s_{sb} * (1-g_{sb}) * sb(t) + (1-g_i) * v_s * mf(t). \tag{4.1.4}$$

The survival probability of non-seedlings function, $s_a(z, \theta)$, is given by:

```{=latex}
\[Logit(s_a(z, \theta)) = \beta_{0,s,i} + \beta_{s,z} * z + \tag{4.1.5} \]
\[\beta_{s, \theta_t, dry} * \theta_{t, dry, i} + \beta_{s, \theta_t, wet} * \theta_{t, wet, i} + \]
\[\beta_{s, \theta_p, dry} * \theta_{p, dry, i} + \beta_{s, \theta_p, wet} * \theta_{p, wet, i} + \]
\[\beta_{s, \theta_{s3}, dry} * \theta_{s3, dry, i} + \beta_{s, \theta_{s3}, wet} * \theta_{s3, wet, i} + \]
\[\beta_{s, \theta_t \times z, dry} * \theta_{t, dry, i} * z + \beta_{s, \theta_t \times z, wet} * \theta_{t, wet, i} * z + \]
\[\beta_{s, \theta_p \times z, dry} * \theta_{p, dry, i} * z + \beta_{s, \theta_p \times z, wet} * \theta_{p, wet, i} * z + \]
\[\beta_{s, \theta_{s3} \times z, dry} * \theta_{s3, dry, i} * z + \beta_{s, \theta_{s3} \times z, wet} * \theta_{s3, wet, i} * z + \] 
\[\beta_{s, native} * g(i) + \beta_{s, i}, \]

```


where $wet$ and $dry$ denote wet season and dry seasons covariation values, $i$ indexs each site in Table 4.1, and the function $g()$ takes a site $i$ and returns 0 for sites in the invaded range and 1 for sites in the native range. The development function, $G(z'|z, \sigma, \theta)$ is given by:

$$G(z'|z, \sigma, \theta) = f_G(z'|\mu_G(z, \theta), \sigma_G(z, i)), \tag{4.1.6}$$

where $f_G$ denotes a normal probability density function, $\mu_G(z,\theta)$ is given by:

```{=latex}

\[\mu_G(z, \theta) = \beta_{0,G,i} + \tag{4.1.7} \]
\[\beta_{G, \theta_t, dry} * \theta_{t, dry, i} + \beta_{G, \theta_t, wet} * \theta_{t, wet, i} + \]
\[\beta_{G, \theta_p, dry} * \theta_{p, dry, i} + \beta_{G, \theta_p, wet} * \theta_{p, wet, i} + \]
\[\beta_{G, \theta_{s3}, dry} * \theta_{s3, dry, i} + \beta_{G, \theta_{s3}, wet} * \theta_{s3, wet, i} + \]
\[\beta_{G, \theta_t \times z, dry} * \theta_{t, dry, i} * z + \beta_{G, \theta_t \times z, wet} * \theta_{t, wet, i} * z + \]
\[\beta_{G, \theta_p \times z, dry} * \theta_{p, dry, i} * z + \beta_{G, \theta_p \times z, wet} * \theta_{p, wet, i} * z + \]
\[\beta_{G, \theta_{s3} \times z, dry} * \theta_{s3, dry, i} * z + \beta_{G, \theta_{s3} \times z, wet} * \theta_{s3, wet, i} * z,\]
\[\beta_{G, native} * g(i) + \beta_{G, i}, \]
```

$\sigma_G(z, i)$ is given by:

```{=latex}
\[\sigma_G(z, \theta) = \beta_{0, \sigma_G, i} + \tag{4.1.8} \]
\[\beta_{\sigma_G, \theta_t, dry} * \theta_{t, dry, i} + \beta_{\sigma_G, \theta_t, wet} * \theta_{t, wet, i} + \]
\[\beta_{\sigma_G, \theta_p, dry} * \theta_{p, dry, i} + \beta_{\sigma_G, \theta_p, wet} * \theta_{p, wet, i} + \]
\[\beta_{\sigma_G, \theta_{s3}, dry} * \theta_{s3, dry, i} + \beta_{\sigma_G, \theta_{s3}, wet} * \theta_{s3, wet, i} + \]
\[\beta_{\sigma_G, native} * g(i) + \beta_{\sigma_G, i}, \]

```

The probability of flowering function, $p_f(z,\theta)$, is given by: 

```{=latex}

\[Logit(p_f(z, \theta)) = \beta_{0, p_f, i} + \beta_{z, p_f} * z \tag{4.1.9} +\]
\[\beta_{p_f, \theta_t, dry} * \theta_{t, dry, i} + \beta_{p_f, \theta_t, wet} * \theta_{t, wet, i} + \]
\[\beta_{p_f, \theta_p, dry} * \theta_{p, dry, i} + \beta_{p_f, \theta_p, wet} * \theta_{p, wet, i} + \]
\[\beta_{p_f, \theta_{s1}, dry} * \theta_{s1, dry, i} + \beta_{p_f, \theta_{s1}, wet} * \theta_{s1, wet, i} + \]
\[\beta_{p_f, \theta_t \times z, dry} * \theta_{t, dry, i} * z + \beta_{p_f, \theta_t \times z, wet} * \theta_{t, wet, i} * z + \]
\[\beta_{p_f, \theta_p \times z, dry} * \theta_{p, dry, i} * z + \beta_{p_f, \theta_p \times z, wet} * \theta_{p, wet, i} * z + \]
\[\beta_{p_f, \theta_{s1} \times z, dry} * \theta_{s1, dry, i} * z + \beta_{p_f, \theta_{s1} \times z, wet} * \theta_{s1, wet, i} * z,\]
\[\beta_{p_f, native} * g(i) + \beta_{p_f, native \times z} * g(i) * z + \beta_{p_f, i}, \]

```

The number of flowers produced conditional on flowering function, $r_f(z,\theta)$, is given by: 

```{=latex}

\[Log(r_f(z, \theta)) = \beta_{0, r_f, i} + \beta_{z, r_f} * z + \tag{4.1.10} \]
\[\beta_{r_f, \theta_t, mean} * \theta_{t, mean, i} + \beta_{r_f, \theta_t, seas} * \theta_{t, seas, i} +  \] 
\[\beta_{r_f, \theta_p, total} * \theta_{p, total, i} + \beta_{r_f, \theta_p, seas} * \theta_{p, seas, i} + \]
\[\beta_{r_f, \theta_{s3}, mean} * \theta_{s3, mean, i} + \beta_{r_f, \theta_{s3}, seas} * \theta_{s3, seas, i} + \]
\[\beta_{r_f, \theta_t \times z, mean} + \theta_{t, mean, i} * z + \beta_{r_f, \theta_t \times z, seas} * \theta_{t, seas, i} * z + \]
\[\beta_{r_f, \theta_p \times z, total} + \theta_{p, total, i} * z + \beta_{r_f, \theta_p \times z, seas} * \theta_{p, seas, i} * z + \]
\[\beta_{r_f, \theta_{s3} \times z, mean} + \theta_{s3, mean, i} * z + \beta_{r_f, \theta_s3 \times z, seas} * \theta_{s3, seas, i} * z + \]
\[\beta_{r_f, native} * g(i) + \beta_{r_f, native \times z} * g(i) * z.\]

```

$g(i)$ is a function that returns 1 if site $i$ in the native range (South Africa) and 0 when site $i$ is located elsewhere.  Finally, the size distribution of newly observed non-seedling plants, $r_a(z')$, is given by:

$$r_a(z') = f_{r_a}(z'| \mu_{r_a}, \sigma_{r_a}), \tag{4.1.11}$$

where $f_{r_a}$ is a Gaussian probability density function.

# Vital rate model summaries

```{r echo = FALSE, message = FALSE, warning = FALSE}

all_surv_mods <- readRDS("../Data/ramet_survival_list_krig.rds")
all_flow_mods <- readRDS("../Data/ramet_flower_n_list_lin_slope_ran.rds")
all_grow_mods <- readRDS("../Data/ramet_growth_list_krig.rds")
all_repr_mods <- readRDS("../Data/ramet_repro_list_krig.rds")


surv_mod <- all_surv_mods[[1]][["times_sw3_seas"]]
flow_mod <- all_flow_mods[[1]][["times_sw3_seas"]]
grow_mod <- all_grow_mods[[1]][["times_sw3_seas"]]
repr_mod <- all_repr_mods[[1]][["times_sw1_seas"]]

recr_mod <- readRDS("../Data/recr_size_brm.rds")


```

\newpage
\blandscape

## Survival model

```{r}

summary(surv_mod)

```

## Growth model

```{r}

summary(grow_mod)

```

## Pr(Flowering) model

```{r}

summary(repr_mod)

```

## Flower number model

```{r}

summary(flow_mod)

```

## Recruit size model

```{r}

summary(recr_mod)

```

\elandscape
\newpage

# Sensitivity of site-level $\lambda$s to simulated parameter values

We could not find parameter values for establishment probability ($p_e$) or seedbank survival rate ($s_{sb}$). Therefore, we simulated a range of values  from 0-1 for each (incrementing by 0.05), re-building the model, and then computing $\lambda$. The results are reported here. Dotted, red vertical lines how the parameter value we used in the results reported in the main text.

## $p_e$

\includepdf[pages = -, nup = 1, pagecommand = {}]{../Figures/p_e_sensitivity_plot_prob_resamp.pdf}

## $s_{sb}$

\includepdf[pages = -, nup = 1, pagecommand = {}]{../Figures/s_sb_sensitivity_plot_prob_resamp.pdf}

